Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-31 Multiresolution_framework/ltx/manuscript.tex
added appendix for smoother

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\Theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\Theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\Theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\Theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in statespace model the E-step is equiavalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\Theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\Theta,\boldsymbol\Theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \Theta'}\left[ .\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \Theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\Theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood.

>>>>

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\Theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\Theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\Theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\Theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in statespace model the E-step is equiavalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\Theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\Theta,\boldsymbol\Theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
where $ \mathbf E_{\boldsymbol \Theta'}\left[ .\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \Theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\Theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood. 
\section{Calculating $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.
The expression for  $\Xi_{0}$ as defined in equation~\ref{eq:defofXi0} of the main text is given by
\begin{equation}\label{eq:Xi0}
 \Xi_1=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right].
We start by expanding the covariance matrix as following
 \mathbf P_{t}&=\mathbf E_{\theta'}\left[(\mathbf x_{t}-\mathbf{\hat x}_{t})(\mathbf x_{t}-\mathbf{\hat x}_{t})^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]-\mathbf E_{\Theta'}\left[\mathbf x_{t}\mathbf{\hat x}_{t}^\top\right]-\mathbf E_{\theta'}\left[\mathbf{\hat x}_{t}\mathbf x_{t}^\top\right]+\mathbf E_{\theta'}\left[\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]-\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top-\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]-\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
therefore we can write
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
substituting \ref{eq:Xi0derivation1} in \ref{eq:Xi0} we get
 \Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
To calculate $\Xi_0$, from equation~\ref{eq:defofXi1} of the main text we have
\begin{equation}\label{eq:Xi1}
 \Xi_0=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
An expression for $\Xi_1$ in terms of the smoothed states can be also found in a similar way by expanding the cross-covariance matrix,
 \mathbf M_{t+1}&=\mathbf E_{\Theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1})^\top\right] \nonumber \\
&=\mathbf E_{\Theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top
therefore we can write
\begin{equation}\label{eq:Xi1derivation1}
 \mathbf E_{\Theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top
substituting \ref{eq:Xi1derivation1} in \ref{eq:Xi1} we get
 \Xi_1=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right)
